---
title: "Land use and land cover classification with satellite image time series in R"
author: "Krzysztof Dyba"
format:
  html:
    toc: true
    toc-title: " "
    embed-resources: true
---

# Introduction

The **sits** is an end-to-end toolkit for analyzing satellite image time
series (SITS) data, primarily focused on land use and land cover (LULC)
classification using machine learning and deep learning models.
It is built on top of two core packages **terra** and **rstac** (and other
spatial packages) and it is designed to handle large-scale data cubes,
including those stored in the cloud environments.

::: {.callout-note appearance="simple"}
**Data cube** is a collection of satellite images covering a specific area
over a period, organized into spatial (x, y) and temporal (time) dimensions,
with spectral bands.
:::

A typical workflow looks like this:

```{mermaid}
flowchart LR
    A[Select image collection<br/>from cloud provider] --> B[Create a regular data cube]
    
    B --> C[Extract time series<br/>from labelled samples]
    
    C --> D[Train a machine<br/>learning model]
    
    D --> E[Classify a data cube and<br/>obtain class probabilities]

    E --> F[Produce classified map]
```

We will start by installing the **sits** package using the `install.packages()`
function. Then, we can load it using the `library()` function.

```{r}
#| message: false

# install.packages("sits")
library("sits")
```

The **sits** package uses the **torch** package to train neural network models,
which requires installing two mandatory libraries, i.e. `LibTorch` and
`LibLantern`. We can do this by calling `torch::install_torch()`. If you
have a slow Internet connection, it is worth increasing the timeout.

```{r}
#| eval: false
#| message: false

options(timeout = 600)
torch::install_torch()
```

# Data acquisition

The first step is to create a data cube. This can be done based on data
available in network services (e.g., Amazon Web Service, Microsoft Planetary
Computer, Copernicus Data Space Ecosystem) or local data. We will choose
the former. To do this, we will use the `sits_cube()` function, in which we
can specify the following arguments: `source`, `collection`, `tiles`, `bands`,
`start_date`, and `end_date`. 

::: {.callout-note appearance="simple"}
To check the available collections, use the `sits_list_collections()` function.
:::

```{r}
cube = sits_cube(
  source = "MPC",
  collection = "SENTINEL-2-L2A",
  tiles = "33UXU",
  bands = c("B02", "B03", "B04", "B08"),
  start_date = "2023-01-01",
  end_date = "2023-12-31",
  progress = FALSE
)
```

The result is a `tibble` (*à la* data frame) object containing metadata about
the data cube and the satellite images themselves. Note that these are just
links to images stored on the server. Naturally, we can display information
about these objects.

```{r}
# data cube
t(cube)
```

```{r}
# images
head(cube$file_info[[1]])
```

In the next step, let's select the appropriate images based on cloud cover
(`cloud_cover` attribute).

```{r}
# select images with less than 10% cloud cover
idx = cube$file_info[[1]]$cloud_cover < 10
cube$file_info[[1]] = cube$file_info[[1]][idx, ]
```

After selection, let's check what imaging dates are available (`date` attribute).

```{r}
unique(cube$file_info[[1]]$date)
```

Using the `plot()` function and defining the relevant spectral bands and date,
we can display the specified scene. For interactive visualization, the
`sits_view()` function can be used.

```{r}
#| message: false

plot(cube, red = "B04", green = "B03", blue = "B02", dates = "2023-05-30")
```

# Time series analysis

```{r}
#| message: false

library("sf")
```

Let's load sample points with specific land cover classes (`points.gpkg` file).
The file is located in the `data` folder. To load, use the `sf` package and
the `read_sf()` function.

```{r}
points = read_sf("data/points.gpkg")
points$category = as.factor(points$category)
```

Let's look at basic information such as the number of points and the coordinate
reference system (CRS).

```{r}
nrow(points)
st_crs(points)$epsg
```

The reference system for points and rasters is identical (`EPSG:32633`), so
we don't need to reproject the data. Let's also check the category distribution
using the `table()` function.

```{r}
table(points$category)
```

We can see that the data is unbalanced (two categories dominate, while the
other three are underrepresented). In practice, this may cause the model to
be unable to learn to recognize these few categories due to the small training
sample.

Let's select two sample objects, for instance cropland and water, and analyze how
their values changed over time. The key function in this case is `sits_get_data()`
which retrieves spectral values from the data cube and creates time series for
specified points. In the function, we must provide the name of the column
(`label_attr` argument) in which the categories are stored.

```{r}
# extract time series for points
idx = c(2, 18) # cropland and water objects
ts_data = sits_get_data(cube, points[idx, ], label_attr = "category")
```

Instead of plotting four spectral curves, let's calculate the NDVI based on
the near-infrared (B08) and red (B04) bands. The `sits_apply()` function is
designed to create new layers, in this case the vegetation index.

```{r}
ts_data = sits_apply(ts_data, NDVI = (B08 - B04) / (B08 + B04))
```

Now let's plot a figure showing how the NDVI changed from April to October
in 2023 for cropland and water. Time series data are stored in the `time_series`
column separately for each point.

```{r}
plot(ts_data$time_series[[1]]$Index, ts_data$time_series[[1]]$NDVI,
     xlab = "Date", ylab = "Value", main = "NDVI", type = "l",
     ylim = c(-0.1, 1), col = "orange")
lines(ts_data$time_series[[2]]$Index, ts_data$time_series[[2]]$NDVI,
     xlab = "Date", col = "blue")
legend("topright", legend = c("Cropland", "Water"), col = c("orange", "blue"),
       lty = 1)
```

# Classification

## Data preparation

Having points with labelled land cover classes and their corresponding spectral
values, we can train a classification model. Developing a model requires training
data and test data (i.e., independent data that was not used for training) to
verify its correctness.

So let's prepare the data. We will split our input dataset (`points` object)
into a training set and a test set in a ratio of 80% and 20%. This procedure
is known as hold-out validation. For reproducibility, set the random seed in
the `set.seed()` function.

```{r}
set.seed(123)

n = round(0.8 * nrow(points))
trainIndex = sample(nrow(points), size = n)
train = points[trainIndex, ]
test = points[-trainIndex, ]
```

Next, to simplify the analysis, we will select only one date from our data
cube, i.e., May 30.

```{r}
scene = cube
idx = which(cube$file_info[[1]]$date == "2023-05-30")
idx = idx[1:4] # remove duplicate scene
scene$file_info[[1]] = cube$file_info[[1]][idx, ]
```

As before, let's extract the spectral values for these two datasets (this may
take a while).

```{r}
train = sits_get_data(scene, train, label_attr = "category")
test = sits_get_data(scene, test, label_attr = "category")
```

In this way, we prepared datasets for modeling.

## Modeling

The **sits** package supports several different machine learning models,
such as *random forest* (`sits_rfor()`), *support vector machines* (`sits_svm()`)
or *gradient boosting* (`sits_xgboost()`), as well as neural network models
(e.g., `sits_mlp()`). To train models, we need the `sits_train()` function,
in which we must define the training data and the model. In our example,
we will use random forest with default hyperparameters without tuning. 

```{r}
rf_model = sits_train(
  samples = train,
  ml_method = sits_rfor()
)
```

## Validation

After training the model, the most important thing we have to do is to verify
its correctness. This is an essential step in modeling that cannot be overlooked.
We need to make a prediction using the trained model (`rf_model` object) on
the test set (`test` object). The `sits_classify()` function is provided for
prediction and will return the predicted labels by the model for each point
based on the already extracted spectral values.

```{r}
test_pred = sits_classify(test, ml_model = rf_model, progress = FALSE)
```

As we know the actual land cover classes and those predicted by the model,
we can use the `sits_accuracy()` function to calculate the accuracy metric,
Kappa coefficient, and confusion (error) matrix. However, this requires the
`caret` package to be installed beforehand.

```{r}
# install.packages("caret")
acc = sits_accuracy(test_pred)
acc[["overall"]][["Accuracy"]]
acc[["overall"]][["Kappa"]]
```

The accuracy value is approximately 0.8. However, it should be noted that
this value is overestimated because our dataset was not balanced. In this case,
the Kappa coefficient is more reliable. Let's also examine the confusion matrix.

```{r}
acc[["table"]]
```

## Prediction

If our model is right, the next step is to make predictions for the entire area.
Since the Sentinel-2 scene is very large and processing will take a long time,
let's limit the analysis area to a smaller extent. As one option, we can define
the extent as four coordinates (`xmin`, `xmax`, `ymin`, `ymax`) in the
corresponding coordinate reference system.

```{r}
roi = c("xmin" = 688000, "xmax" = 695000, "ymin" = 5826000, "ymax" = 5830000)
```

::: {.callout-note appearance="simple"}
Alternatively, it is also possible to use the `sf` object (polygon).
:::

We can download data from the server to the disk using the `sits_cube_copy()`
function. It allows setting the output spatial extent (`roi` argument), so
the images will be cropped to the specified area. We will save the data in a
temporary folder by specifying the `tempdir()` function in the `output_dir`
argument. Of course, we can also provide a different directory.

```{r}
scene = sits_cube_copy(
  scene,
  roi = roi,
  output_dir = tempdir(),
  progress = FALSE
)
```

We have saved the cropped rasters locally in a temporary folder. Now we proceed
to the actual prediction stage, which consists of two steps:

1. Data cube classification and obtaining class probabilities for each raster
cell.
2. Creating a classification map based on the highest probability values.

For the first step, we will need the `sits_classify()` function that we used
before. However, this time, we will use scenes (images) as input data instead
of points.

```{r}
probs_cube = sits_classify(
  data = scene,
  ml_model = rf_model,
  output_dir = tempdir(),
  progress = FALSE
)
```

The second step requires engaging the `sits_label_classification()` function
on the data cube with the calculated class probabilities. In this case, we will
create a new folder called `results` where we will save the classification
result.

```{r}
dir.create("results")
```

```{r}
classification = sits_label_classification(
  probs_cube,
  output_dir = "results",
  progress = FALSE
)
```

## Visualization

Finally, let's see what the classification results look like on the map!
To load the raster, we will use the **terra** package and the `rast()` function.
An important step is to assign appropriate labels (land classes) to the raster
cells.

```{r}
#| message: false

library("terra")

filepath = classification$file_info[[1]]$path
raster = rast(filepath)
```

The IDs and corresponding classes are stored in the `classification$labels`
vector. We need to convert it to a data frame and then assign it to `raster`
using the `levels()` function.

```{r}
df = data.frame(id = as.numeric(names(classification$labels[[1]])),
                category = classification$labels[[1]])
levels(raster) = df
```

With the labels prepared, we can create a map by pre-defining the appropriate
colors.

```{r}
colors = c("#d20000", "#fdd327", "#d9d9d9", "#29a329", "#1445f9")
plot(raster, col = colors, main = "Classified image", mar = c(1, 1, 1, 5.5))
```

::: {.callout-note appearance="simple"}
In pixel-based classification, the salt-and-pepper effect is a common issue
(as seen in the image above). To counteract this, functions such as
`sits_smooth()` in pre-processing or `terra::focal(fun = "modal")` and
`terra::sieve()` in post-processing can be used.
:::

# Summary

In this tutorial, we covered the basics of satellite data analysis in **sits**
package. It provides a comprehensive framework for land use and land
cover classification using satellite image time series data including data cube
management, data processing, model training, and visualization. The package is
actively developed, so check the [GitHub repository](https://github.com/e-sensing/sits)
for the latest features and updates. Definitely check out the [book](https://e-sensing.github.io/sitsbook/)
as well!

# Task

Train another machine learning model based on the `points.gpkg` data. Compare
its accuracy with the trained random forest model. Then, using the better model,
make predictions for two selected dates (e.g., 2016 and 2024). Have there been
any notable changes in land cover?
